---
title: "Two-Source Point Source Model on Chorley Ribble Data"
geometry: "left=2cm,right=2cm,top=1cm,bottom=1cm"
output:
  pdf_document: 
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
Sys.setenv(lang = "en_US")
```

```{r}
library(spatstat)
```

# Adding another source

The first incinerator information is stored in `chorley.extra`. The incinerator is located at (x = 354.5, y = 413.6). We will add another one at different locations. We'll use the same windows as is used for the Chorley data.
Example locations of the additional incinerator:
- (x = 355, y = 414)
- (x = 355.5, y = 414.5)
- (x = 365, y = 415)
- (x = 355, y = 428)

```{r}
sources <- ppp(c(354.5,355.5), c(413.6,414.5), window = chorley$window)
```

```{r}
S <- split(chorley)
larynx <- S$larynx # cases
lung <- S$lung # controls
```

```{r}
X <- layered(larynx, sources)
layerplotargs(X)[[1]] <- list(pch = 4, cex = 0.5)
layerplotargs(X)[[2]] <- list(pch = 20, col = "red")
plot(X, main = "Potential source and larynx cancer cases")
```

```{r}
smo <- density(lung, sigma = 0.15, eps = 0.1, positive = TRUE)
smo <- eval.im(pmax(smo, 1e-10))
```

```{r}
Q <- quadscheme(larynx, eps = 0.1)
```

# Fit Null Model
```{r}
chorley0fit <- ppm(Q ~ offset(log(smo)))
```

```{r}
chorley0fit
```

# Fit Alternative Model

```{r}
ls_all_raisins = list()
for (i in 1:sources$n)  {
    ls_all_raisins[i] =  paste0("(1 + alpha * exp(-beta * ((x- ",sources$x[i],")^2 + (y- ",sources$y[i],")^2)))")
}
```

```{r}
str_all_raisins <- paste(ls_all_raisins, collapse = "*")
```

```{r}
eval(parse(text = paste('raisin_func <- function(x, y, alpha, beta) {(' , str_all_raisins , ')}', sep='')))
```

```{r}
chorleyDfit <- ippm(Q ~ offset(log(smo) + log(raisin_func)),
  start = list(alpha = 5, beta = 1), gcontrol = glm.control(maxit = 1000))
```

```{r}
chorleyDfit
```

# Compare the two models by the generalize likelihood ratio test
```{r}
anova(chorley0fit, chorleyDfit, test = "LRT")
```


