---
title: "Traceback_Model_based_Diggle_1997"
author: "Sandra Rudeloff"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Derivation of the likelihood function

Likelihood function:
$$\mathcal{L}( \theta, \rho) = - \sum_{i=1}^p \mu_i + \sum_{i=1}^p y_i \log{\mu_i}$$

-   $\theta$: risk parameters, in my case $\alpha$ and $\beta$
-   $p$: number of subregions - in my case 100
-   $y_i$: number of cases in each subregion
-   $\mu_i$ expected number of cases in subregion $i$

Expected Number of cases: $$\mu_i = \rho N_i \prod_{l=1}^{k} f_l(x_i)$$

-   $\rho$: scale parameter
-   $N_i$: number of people at risk in subregion $i$, in my case the
    population
-   $f_l(x_i)$: risk function associated with the $l^{th}$ source,
    evaluated at the centroid $x_i$ of subregion $i$.

risk function for distance $d$ from the source:
$$f(d) = 1 + \alpha \exp[- (\frac{d}{\beta})^2]$$

-   $\alpha$: proportional elevation in risk at the source
-   $\beta$: decrease in risk

Given this information, you can use the likelihood function to estimate
the parameters $\alpha, \beta$ and $\rho$ for the model without
covariates.

# Libraries

```{r}
library(readxl)
library(ggplot2)
library(dplyr)
library(ggnewscale)
library(RColorBrewer)
library(plotly)
source("generate_population.R")
```

# Input Data

```{r}
investigation_scenario <- 2
no_of_cells <- 100
```

# 1. Collect Variables

## Population Data

number of people at risk in each subregion $N$

```{r}
read_population_data <- function(investigation_scenario,no_of_cells ) {  
  cells_per_row <- sqrt(no_of_cells)
  
  # generate a sequence of coordinates for centroids and generate all combinations of these coordinates
  centroid_coords <- seq(50, by = 100, length.out = cells_per_row)
  df_population <- expand.grid(x_centroid = centroid_coords, y_centroid = centroid_coords)
  
  population_type <- subset(scenarios_data_population, scenario == investigation_scenario)$population_type
  total_population <- subset(scenarios_data_population, scenario == investigation_scenario)$total_population
  
  if (population_type == "radial_clusters" || population_type ==  "main_and_small_clusters" || population_type == "linear") {
    # used for all radial type populations
    desired_gradient <- subset(scenarios_data_population, scenario == investigation_scenario)$desired_gradient
    # high values mean a large spreading 
  }
  
  if (population_type == "radial_clusters" || population_type ==  "main_and_small_clusters" ) {
    num_clusters <- subset(scenarios_data_population, scenario == investigation_scenario)$num_clusters
  }
  
  # generate population
  df_population <- switch(population_type,
                          "random" = generate_random_population(df_population, total_population),
                          "uniform" = generate_uniform_population(df_population, total_population),
                          "linear" = generate_linear_population(df_population, total_population, desired_gradient),
                          "radial_clusters" = generate_radial_clusters_population(df_population, total_population, desired_gradient, num_clusters),
                          "main_and_small_clusters" = generate_main_and_small_clusters_population(df_population, total_population, desired_gradient, num_clusters)
  )
  
  return(df_population)
}

scenarios_data_population <- read_excel("./Data/scenarios.xlsx", sheet = "Population")

df_population <- read_population_data(investigation_scenario, no_of_cells) 
```

## Outbreak Data
```{r}
df_outbreak <- subset(read_excel("./Data/scenarios.xlsx", sheet = "Outbreak_Locations"), scenario == investigation_scenario)[,2:3]
```

## Shops Data
```{r}
df_shops <- subset(read_excel("./Data/scenarios.xlsx", sheet = "Store_Locations"), scenario == investigation_scenario)[,2:4]
```

## Visualization 
Assign colors to different chains
```{r}
# Identify unique chains and generate a color palette
unique_chains <- unique(df_shops$chain)
num_chains <- length(unique_chains)

# Generate a color palette avoiding red and orange
all_colors <- brewer.pal(9, "Set1")  # This palette has distinct colors
all_colors <- all_colors[!all_colors %in% c("#E41A1C", "#377EB8")] 
chain_colors <- all_colors[1:num_chains]

# Create a named vector for chain colors
names(chain_colors) <- unique_chains
```

Assign breaks for the legend of the population 
```{r}
min_population <- min(df_population$population)
max_population <- max(df_population$population)

# Calculate the range of population values
range_population <- max_population - min_population

# Dynamically determine the step size based on the range
if (range_population <= 10) {
  step_size <- 2
} else if (range_population <= 20) {
  step_size <- 5
} else if (range_population <= 50) {
  step_size <- 10
} else {
  step_size <- 15
}

# Generate the sequence of breaks
breaks_population <- seq(min_population, max_population, by = step_size)
```

Visualize
```{r}
ggplot() +
  # Plot the population data
  geom_tile(data=df_population, aes(x=x_centroid, y=y_centroid, fill=population), width = 100, height = 100, alpha = 0.8) +
  scale_fill_gradient(low = "white", high = "cadetblue",name = "Population", breaks = breaks_population) +
  

  # Introduce a new scale for the shop chains
  new_scale_fill() +
  
  # Plot the shops data
  geom_point(data=df_shops, aes(x=store_x, y=store_y, fill=chain), size=3, shape=23, alpha = 0.8) +
  scale_fill_manual(values=chain_colors, name="Shop Chain") +
  
  # Plot the outbreak data
  geom_point(data=df_outbreak, aes(x=case_x, y=case_y), color="red", size=2, shape=21, fill="red", alpha=0.8) +

  # Adjust the x and y axis breaks to have lines every 100m
  scale_x_continuous(breaks=seq(0, 1000, by=100)) +
  scale_y_continuous(breaks=seq(0, 1000, by=100)) +
  
  # Add labels and theme
  labs(title=sprintf("Visualization of Scenario %s", investigation_scenario),
       x="X Coordinate",
       y="Y Coordinate",
       color="Shop Chain") +
  theme_minimal() +
  theme(legend.position="bottom", aspect.ratio=1,  panel.grid.minor = element_blank()) 
```




# 2. Model
```{r}
# 2. Calculate the Risk Function for Each Source and Subregion ---- 
# For each subregion centroid, compute the risk function for each store (source) based on the distance between the subregion centroid and the store.

# Risk function for distance d from the source
risk_function <- function(d, alpha, beta) {
  return(1 + alpha * exp(- (d/beta)^2))
}

# Compute distance between two points
compute_distance <- function(x1, y1, x2, y2) {
  return(sqrt((x1 - x2)^2 + (y1 - y2)^2))
}

# Compute the risk matrix for each subregion and source
compute_risk_matrix <- function(df_population, df_shops, alpha, beta) {
  risk_matrix <- matrix(0, nrow(df_population), nrow(df_shops))
  for (i in 1:nrow(df_population)) {
    for (j in 1:nrow(df_shops)) {
      d <- compute_distance(df_population$x_centroid[i], df_population$y_centroid[i], df_shops$store_x[j], df_shops$store_y[j])
      risk_matrix[i, j] <- risk_function(d, alpha, beta)
    }
  }
  return(risk_matrix)
}

# 3. Compute Expected Number of Cases ----
#For each subregion, compute the expected number of cases (mu) using the risk functions and the population at risk in that subregion.

# 4. Define the Likelihood Function ----
# Using the expected number of cases and the observed number of cases, define the likelihood function.
# Likelihood function
likelihood_function <- function(params, y, N) {
  alpha <- params[1]
  beta <- params[2]
  rho <- params[3]
  print(paste("alpha: ", alpha, "beta: ", beta, "rho: ", rho))
  
  risk_matrix <- compute_risk_matrix(df_population, df_shops, alpha, beta)
  mu <- rho * N * apply(risk_matrix, 1, prod)

  L <- sum(-mu + y * log(mu))
  print(paste("likelihood: ", L))
  
  return(-L) # We return the negative likelihood because optimizers in R typically minimize
}

# Construct Y
# Define a function to check if a point is within a square centered at (x, y) with side length 2*delta
point_in_square <- function(px, py, x, y, delta) {
  return(px >= (x - delta) & px <= (x + delta) & py >= (y - delta) & py <= (y + delta))
}

# Define a delta (half the side length of the square)
delta <- 50  # You can adjust this value based on your data's scale

# Observed number of cases in each subregion
y <- numeric(nrow(df_population))
for (i in 1:nrow(df_population)) {
  y[i] <- sum(point_in_square(df_outbreak$case_x, df_outbreak$case_y, df_population$x_centroid[i], df_population$y_centroid[i], delta))
}

# 5. Optimize the Likelihood Function ----
# Use an optimization algorithm to estimate the parameters (alpha, beta, and rho)
# Number of people at risk in each subregion
N <- df_population$population

# Initial parameter values
initial_params <- c(alpha=100, beta=100, rho=1)

# Optimization
# Lower bounds for the parameters
lower_bounds <- c(alpha=0, beta=0, rho=0.00000001)

# Optimization with constraints
result <- optim(par=initial_params, fn=likelihood_function, y=y, N=N, method="L-BFGS-B", lower=lower_bounds, hessian = TRUE)

# Extract estimated parameters
estimated_params <- result$par
print(estimated_params)
```


```{r}
lower_bounds <- c(alpha=0, beta=0, rho=0)
upper_bounds <- c(alpha=1000, beta=1000, rho=10)  # You can adjust these bounds based on your domain knowledge.
result <- DEoptim(fn=likelihood_function, lower=lower_bounds, upper=upper_bounds, y=y, N=N)
estimated_params <- result$optim$bestmem
print(estimated_params)



# Fit the null model
null_likelihood_function <- function(params, y, N) {
  rho <- params[1]
  mu <- rho * N
  
  L <- sum(-mu + y * log(mu))
  return(-L)
}

# Optimize the null model
initial_params_null <- c(rho=1)
result_null <- optim(par=initial_params_null, fn=null_likelihood_function, y=y, N=N, method="SANN", hessian = TRUE)
logLik_null <- -result_null$value

# Optimize the alternative model using DEoptim

lower_bounds <- c(alpha=0, beta=0, rho=0)
upper_bounds <- c(alpha=1, beta=1, rho=10)
result_alternative <- DEoptim(fn=likelihood_function, lower=lower_bounds, upper=upper_bounds, y=y, N=N)
logLik_alternative <- -result_alternative$optim$bestval
```
# Compute Standard errors 

# Compare the nested Models
```{r}
# Compute the GLRT statistic
GLRT_statistic <- -2 * (logLik_null - logLik_alternative)

# Determine the degrees of freedom (difference in number of parameters between the two models)
df <- 2  # alpha and beta are the additional parameters in the alternative model

# Compute the p-value
p_value <- 1 - pchisq(GLRT_statistic, df)

# Print the results
print(paste("GLRT statistic:", GLRT_statistic, "\n"))
print(paste("Degrees of freedom:", df, "\n"))
print(paste("P-value:", p_value, "\n"))

# Decide on the hypothesis based on a significance level (e.g., 0.05)
if (p_value < 0.05) {
  cat("Reject the null hypothesis in favor of the alternative.\n")
} else {
  cat("Fail to reject the null hypothesis.\n")
}


```


